Linting Linearity in Core/System FC

### 2023-01-18 report.tex
Write more on the length-indexed list example

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
higher-kinded\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
constructor}} type \textbf{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a wildly popular type level
feature that allow the user to define data types by enumerating its constructors
with explicit type signatures, where the result type may differ in the type
parameters across branches; pattern matching against GADTs might introduce type
refinements locally, i.e. in individual case alternatives. We develop the
length-indexed lists example without discussing the type system and type
inference details of GADTs which we explore later in~\ref{related-work-gadts}.
First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and}, the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist due to the
\emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.
%
\begin{code}
    data Nat = Z | S Nat
\end{code}
% Before introducing the data type of

Besides the aforementioned type-level features of Haskell, \emph{linear types},
which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system that makes
it yet more expressive.

>>>>

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
non-nullary\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
constructor}} type constructor \textbf{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
% include recursive ADTs
\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
which users can define data types as they would common recursive algebraic data
types, but that allows explicit type signatures for the data constructors where
the result type may differ in the type parameters across branches. Pattern
matching against GADTs can introduce local type refinements, that is, pattern
matching against certain data constructor will refine the type information known
for typechecking in individual case alternatives. We develop the length-indexed
lists example without discussing the type system and type inference details of
GADTs which we explore later in~\ref{related-work-gadts}.

% TODO: Explain Type in Type?

First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist because of
\emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.

\begin{code}
    data Nat = Z | S Nat
\end{code}

Secondly, we define the data type in GADT syntax for length-index lists which
takes two type parameters. The first type parameter is the length of the list
and the type of the type parameter (i.e. the kind of the first type parameter)
is \textbf{Nat}. To construct a type of kind \textbf{Nat} we can only use the
type constructors \textbf{Z} and \textbf{S}. The second type parameter is the
type of the values contained in the list, and any type is valid, hence the
\textbf{Type} kind.

\begin{code}
    data Vec (n :: Nat) (a :: Type) where
        Nil :: Vec Z a
        Cons :: a -> Vec m a -> Vec (S m) a
\end{code}

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
possible different data type constructors being parametrized over different type
parameters as we do with \textbf{Vec}'s size parameter being different depending
on the constructor that constructs the list.
    To define the safe \textbf{head} function, we must specify the type of the
input list taking into account that the size must not be zero. To that effect,
the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.

\begin{code}
    head :: Vec (S n) a -> a
    head (Cons x xs) = x
\end{code}

Pattern matching on the \textbf{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\textbf{Nil} -- it would entail the list has size zero; however, it's impossible
to call \textbf{head} on a list of size zero!

Linear types, which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system, besides the aforementioned type-level
features, that makes it even more expressive.
